Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-17 output/latex/manuscript.tex
rename section "Methods and materials" to "Methods"
update latex

%% pandoc-fignos: environment for tagged figures
\newenvironment{fignos:tagged-figure}[1][]{
  \let\oldthefigure\thefigure
  \let\oldtheHfigure\theHfigure
  \renewcommand{\thefigure}{#1}
  \renewcommand{\theHfigure}{#1}
}{
  \let\thefigure\oldthefigure
  \let\theHfigure\oldtheHfigure
  \addtocounter{figure}{-1}
}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms},
  pdfauthor={Milton PividoriSumei LuBinglan LiChun SuMatthew E. JohnsonWei-Qi WeiQiping FengBahram NamjouKrzysztof KirylukIftikhar KulloYuan LuoBlair D. SullivanBenjamin F. VoightCarsten SkarkeMarylyn D. RitchieStruan F.A. GrantCasey S. Greene},
  pdflang={en-US},
  pdfkeywords={genetic studies, functional genomics, gene co-expression, therapeutic targets, drug repurposing, clustering of complex traits},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The cardiovascular sub-branch had 129 significant LV-trait associations in PhenomeXcan and 23 in eMERGE.
LV136, aligned with known collagen formation and muscle contraction pathways (Supplementary Table \ref{tbl:sup:multiplier_pathways:lv136}), was associated with coronary artery disease and keratometry measurements in PhenomeXcan (Supplementary Tables \ref{tbl:sup:phenomexcan_assocs:lv136}).
In eMERGE, this LV was associated with coronary atherosclerosis (phecode: 411.4) (Supplementary Table \ref{tbl:sup:emerge_assocs:lv136}).
LV136 was expressed in a wide range of cell types, including fibroblasts, mesenchymal stem cells, osteoblasts, pancreatic stellate cells, cardiomyocytes, and adipocytes (Supplementary Figure \ref{fig:sup:lv136}).
Within the cardiovascular sub-branch, we found neuropsychiatric and neurodevelopmental disorders such as Alzheimer's disease, schizophrenia, and attention deficit hyperactivity disorder (ADHD).
These disorders were previously linked to the cardiovascular system\textsuperscript{\protect\hyperlink{ref-10qjLoufR}{83}--\protect\hyperlink{ref-13t4TuFeJ}{86}} and share several risk factors, including hypertension, high cholesterol, obesity, smoking, among others\textsuperscript{\protect\hyperlink{ref-D83Aqhga}{87},\protect\hyperlink{ref-9BGyO071}{88}}.
However, our results grouped these diseases by potentially shared transcriptional processes expressed in specific tissues/cell types.
Alzheimer's disease (not present in eMERGE), for instance, was significantly associated with LV21 in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv21}).
LV21, a gene module not aligned to prior pathways, was strongly expressed in a variety of soft tissue sarcomas, monocytes/macrophages (including microglia from cortex samples), and aortic valves (Supplementary Figure \ref{fig:sup:lv21} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv21}).
This LV was also strongly associated with lipids and high cholesterol in PhenomeXcan and hyperlipidemia (phecode: 272.1) in eMERGE (Supplementary Table \ref{tbl:sup:emerge_assocs:lv21}).
As discussed previously, macrophages play a key role in the reverse cholesterol transport and thus atherogenesis\textsuperscript{\protect\hyperlink{ref-idlBgtFz}{89}}, and lipid metabolism in microglia has been recently identified as an important factor in the development of neurodegenerative diseases\textsuperscript{\protect\hyperlink{ref-18I4ish9s}{90}}.

@@

\hypertarget{sec:methods}{%
\subsection{Methods and materials}\label{sec:methods}}

@@

\begin{fignos:tagged-figure}[S1]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S2]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S3]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S4]

@@

\end{fignos:tagged-figure}

\begin{fignos:tagged-figure}[S5]

@@

\end{fignos:tagged-figure}

>>>>

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms},
  pdfauthor={Milton PividoriSumei LuBinglan LiChun SuMatthew E. JohnsonWei-Qi WeiQiping FengBahram NamjouKrzysztof KirylukIftikhar KulloYuan LuoBlair D. SullivanBenjamin F. VoightCarsten SkarkeMarylyn D. RitchieStruan F.A. GrantCasey S. Greene},
  pdflang={en-US},
  pdfkeywords={genetic studies, functional genomics, gene co-expression, therapeutic targets, drug repurposing, clustering of complex traits},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The cardiovascular sub-branch had 129 significant LV-trait associations in PhenomeXcan and 23 in eMERGE.
LV136, aligned with known collagen formation and muscle contraction pathways (Supplementary Table \ref{tbl:sup:multiplier_pathways:lv136}), was associated with coronary artery disease and keratometry measurements in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv136}).
In eMERGE, this LV was associated with coronary atherosclerosis (phecode: 411.4) (Supplementary Table \ref{tbl:sup:emerge_assocs:lv136}).
LV136 was expressed in a wide range of cell types, including fibroblasts, mesenchymal stem cells, osteoblasts, pancreatic stellate cells, cardiomyocytes, and adipocytes (Supplementary Figure \ref{fig:sup:lv136}).
Within the cardiovascular sub-branch, we found neuropsychiatric and neurodevelopmental disorders such as Alzheimer's disease, schizophrenia, and attention deficit hyperactivity disorder (ADHD).
These disorders were previously linked to the cardiovascular system\textsuperscript{\protect\hyperlink{ref-10qjLoufR}{83}--\protect\hyperlink{ref-13t4TuFeJ}{86}} and share several risk factors, including hypertension, high cholesterol, obesity, smoking, among others\textsuperscript{\protect\hyperlink{ref-D83Aqhga}{87},\protect\hyperlink{ref-9BGyO071}{88}}.
However, our results grouped these diseases by potentially shared transcriptional processes expressed in specific tissues/cell types.
Alzheimer's disease (not present in eMERGE), for instance, was significantly associated with LV21 in PhenomeXcan (Supplementary Table \ref{tbl:sup:phenomexcan_assocs:lv21}).
LV21, a gene module not aligned to prior pathways, was strongly expressed in a variety of soft tissue sarcomas, monocytes/macrophages (including microglia from cortex samples), and aortic valves (Supplementary Figure \ref{fig:sup:lv21} and Supplementary Table \ref{tbl:sup:multiplier_pathways:lv21}).
This LV was also strongly associated with lipids and high cholesterol in PhenomeXcan and hyperlipidemia (phecode: 272.1) in eMERGE (Supplementary Table \ref{tbl:sup:emerge_assocs:lv21}).
As discussed previously, macrophages play a key role in the reverse cholesterol transport and thus atherogenesis\textsuperscript{\protect\hyperlink{ref-idlBgtFz}{89}}, and lipid metabolism in microglia has been recently identified as an important factor in the development of neurodegenerative diseases\textsuperscript{\protect\hyperlink{ref-18I4ish9s}{90}}.

@@

\hypertarget{sec:methods}{%
\subsection{Methods}\label{sec:methods}}

@@

@@

@@

@@

@@

@@
